A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-09-07 bioc-workflow.Rmd
added data and updated manuscript

Data quality is routinely examined through visualisation to verify
that sub-cellular niches have been separated along the gradient. Based
on De Duve's principle [@DeDuve:1981] proteins that co-localise
exhibit similar quantitation profiles across the gradient fractions
employed. One approach that has been widely used to visualise and
inspect high throughput mass spectrometry-based proteomics data is
principal components analysis (PCA). PCA is one of many dimensionality
reduction methods, that allow one to effectively summarise
multi-dimensional data in to 2 or 3 dimensions to enable
visualisation. Very generally, the original continuous
multi-dimensional data is transformed into a set of orthogonal
components ordered according to the amount of variability that they
describe. The `plot2D` method in `r Biocpkg("pRoloc")` allows one to
plot the principal components (PCs) of a dataset against one another,
by default the first two components are plotted on the x- and y-axis,
respectively (the `dims` argument can be used to plot other PCs). If
distinct clusters are observed, we assume that there is organellar
separation present in the data. In the code chunk below we produce a
PCA plot of the mouse stem cell dataset. One point on the plot
represents one protein. We can indeed see several distinct protein
clusters. Although, representing the multi-dimensional data along a
limited set of PCs does not give us a hard quantitative measure of
separation, it is extremely useful summarising complex experimental
information in one figure, to get an simplified overview of the data.

In the code chunk below, we specify `fcol = NULL`, which means not to
consider any feature variable to annotate the features (proteins) with
colours. We will see later how to use this to annotate the PCA plot
with prior information about sub-cellular localisation.

@@

These markers can then be mapped to an `MSnSet`'s `featureNames`. The
mouse dataset used here has Uniprot IDs stored as the `featureNames`
(see `head(featureNames(lopit2016))`) and the names of the vector of
the mouse markers stored in `r Biocpkg("pRoloc")` (`mmus` markers) 
are also Uniprot IDs (see `head(mrk)` in the code chunk below), so 
it is straightforward to match names between the markers and 
the `MSnSet` instance using the `addMarkers` function.

@@

If the naming between the marker sets and the `MSnSet` dataset are
different, one will have to convert and match the proteins according
to the appropriate identifier. Sometimes, we find the equivalent entry
name, Uniprot ID or accession number is stored with the data, which
makes conversion between identifers relatively straightforward. If
this is not the case however, conversion can be performed using the 
`r Biocpkg("biomaRt")`, the Bioconductor
[annotation resouces](http://bioconductor.org/help/workflows/annotation/Annotation_Resources/)
or any conversion softwares available online.

We now visualise these annotations along the PCA plot using the
`plot2D` function. We also use the `addLegend` function to map the
marker classes to the pre-defined colours. We also display the data
along the first and seventh PCs using the `dims` argument. Note that
in this second call to the `plot2D` function, we have omitted the
`fcol` argument, as the `"markers"` feature variable name is the
default value. We choose to display PCs 1 and 7 to illustrate that
while upper principal components explain much less variability in the
data (2.23% for PC7, as opposed to 48.41% for PC1), we see that the
mitochondrial and peroxisome clusters can be differenciated, despite
the apparent overlap in the two first PCs.

@@

```{r}
hl <- fDataToUnknown(hl, from = "Golgi", to = "unknown")
getMarkers(hl)
```

@@

```{r pdtab, echo=FALSE, eval = FALSE}
i <- c(2, 3, 4, 13, 14)
pData(hl)[i, -3]
```

@@

In addition to extracting new phenotypes, novelty detection methods 
are also useful for confirming the presence of known or postulated 
clusters in an unbiased fashion. For example, in [@hyper] 
the `phenoDisco` algorithm was used to confirm the data-specific
presence of the nucleus and nucleus sub-compartments. In the code 
chunk below, we re-run this analysis, highlighting some of the 
optional arguments and parameters available for phenotype extraction 
and give some advice on how to interpret the output.

As the `phenoDisco` algorithm is semi-supervised it uses both 
labelled (markers) and unlabelled data to explore the data structure 
and find new sub-cellular data clusters. Thus the first step is to 
define some input labelled data i.e. markers, that the algorithm 
will use as input for the supervised learning aspect of the 
algorithm. As described in [@hyper] we define a set of markers 
to use as input for the analyses that cover well-known residents 
from three distinct organelle structures; the mitochondria, plasma 
membrane and ER, and from three well-known and abundant protein 
complexes; the proteasome and two ribosomal subunits, 40S and 60S. 
These input markers are stored in the `featureData` column of 
`hl` where `fcol = "phenoDisoc.Input"`. We can use the convenience 
accessor function `getMarkers` to print out a table of the markers 
contained in this marker set. These initial markers were manually 
curated using information from the UniProt database, the Gene 
Ontology and the literature.

@@

In the code chunk below we run the `phenoDisco` function and 
return a novelty detection result stored in `pdRes`, according 
to the specified parameters. The algorithm parameters `times`, 
`GS` and `p` are passed to the function, along with the `fcol` 
telling the algorithm where the input training data is contained.

```{r runPD, eval = FALSE}
## As per Christoforou et al (2016), 
hl <- phenoDisco(hl, fcol = "phenoDisco.Input", 
                 times = 200, 
                 GS = 60)

The argument `times` defined the number of times we run unsupervied 
Gaussian Mixture Modelling before defining a new phenotype cluster. 
The recommended minimum and default value is 100. In the below 
code chunk we increase the value to `times = 200` as we have found 
for larger datasets (e.g. 5000+ proteins) a higher times is requried 
for convergence. `GS` defines the minimum number of proteins allowed 
per new data cluster and thus heavily influences what type of new 
clusters are extracted. For example, if a user is interesed in the 
detection of small complexes they may wish to use a small `GS = 10`, 
or `=20` etc. If they wish to detect larger more abundant sub-cellular 
niches a higher `GS` would be preferable. Specifying a small `GS` is be 
more time consuming than a larger `GS`, and there is a trade off 
between finding interesting small complexes and those that may not 
be of interest as we find there is a tendancy to find more noise when 
using a small `GS` compared to using a higher one.

One may also consider increasing the search space for new data 
clusters by increasing the value of the parameter `G`. This defines 
the number of GMM components to test and fit; the default is `G = 1:9` 
(`mclust`'s default). One should note that the choice of decreasing 
`GS`, and increasing the values of the arguments `times`, `G` 
(among other function arguments, see `?phenoDisco`) will heavily 
influence (increase) the total time taken to run the algorithm. 
`phenoDisco` supports parallelisation and we strongly suggest 
you make use of a parallel processing to run these analyses (different 
backends can be set with argument `BPPARAM`).

The ouput of running the `phenoDisco` algorithm is a `MSnSet` containing 
the new data clusters, appended to the `featureData` under the name 
`pd`. We can see by typing `pdRes` directly into the console the 
processing information has been updated to the `MSnSet` recording the 
parameters that were used to run the analyses. This is handy for keeping 
track of data analyses. The results can be displayed by using the 
`getMarkers` function. We see that 5 new phenotype data clusters were 
found.

```{r loadPDres, include=FALSE}
load("Data/pd_runs/pd.rda")
hl <- pdRes
setStockcol(NULL)
setStockcol(paste0(getStockcol(), 70))
```
```{r showPDres}
hl
getMarkers(hl, fcol = "pd")
```

@@

We can also examine the each phenotype intercatively and visualise 
their protein profiles by using the `pRolocVis` function in the 
`pRolocGUI`.

>>>>

Data quality is routinely examined through visualisation to verify
that sub-cellular niches have been separated along the gradient. Based
on De Duve's principle [@DeDuve:1981] proteins that co-localise
exhibit similar quantitation profiles across the gradient fractions
employed. One approach that has been widely used to visualise and
inspect high throughput mass spectrometry-based proteomics data is
principal components analysis (PCA). PCA is one of many dimensionality
reduction methods, that allow one to effectively summarise
multi-dimensional data in to 2 or 3 dimensions to enable
visualisation. Very generally, the original continuous
multi-dimensional data is transformed into a set of orthogonal
components ordered according to the amount of variability that they
describe. The `plot2D` method in `r Biocpkg("pRoloc")` allows one to
plot the principal components (PCs) of a dataset against one another,
by default the first two components are plotted on the x- and y-axis,
respectively (the `dims` argument can be used to plot other PCs). If
distinct clusters are observed, we assume that there is organellar
separation present in the data. Although, representing the 
multi-dimensional data along a
limited set of PCs does not give us a hard quantitative measure of
separation, it is extremely useful summarising complex experimental
information in one figure, to get an simplified overview of the data.

In the code chunk below we produce a
PCA plot of the mouse stem cell dataset. One point on the plot
represents one protein. We can indeed see several distinct protein
clusters. We specify `fcol = NULL`, which means not to
consider any feature variable to annotate the features (proteins) with
colours. We will see later how to use this to annotate the PCA plot
with prior information about sub-cellular localisation.

@@

These markers can then be mapped to a `MSnSet`'s `featureNames`. The
mouse dataset used here has Uniprot IDs stored as the `featureNames`
(see `head(featureNames(hl))`) and the names of the vector of
the mouse markers stored in `r Biocpkg("pRoloc")` (`mmus` markers) 
are also Uniprot IDs (see `head(mrk)` in the code chunk below), so 
it is straightforward to match names between the markers and 
the `MSnSet` instance using the `addMarkers` function.

@@

If the naming between the marker sets and the `MSnSet` dataset are
different, one will have to convert and match the proteins according
to the appropriate identifier. Sometimes, we find the equivalent entry
name, Uniprot ID or accession number is stored with the data, which
makes conversion between identifers relatively straightforward. If
this is not the case however, conversion can be performed using 
`r Biocpkg("biomaRt")`, the Bioconductor
[annotation resouces](http://bioconductor.org/help/workflows/annotation/Annotation_Resources/)
or any conversion softwares available online.

We now visualise these annotations along the PCA plot using the
`plot2D` function and then use the `addLegend` function to map the
marker classes to the pre-defined colours. We also display the data
along the first and seventh PCs using the `dims` argument. Note that
in this second call to the `plot2D` function, we have omitted the
`fcol` argument, as the `"markers"` feature variable name is the
default value. We choose to display PCs 1 and 7 to illustrate that
while upper principal components explain much less variability in the
data (2.23% for PC7, as opposed to 48.41% for PC1), we see that the
mitochondrial and peroxisome clusters can be differenciated, despite
the apparent overlap in the two first PCs.

@@

```{r}
hl <- fDataToUnknown(hl, from = "Golgi apparatus", to = "unknown")
getMarkers(hl)
```

@@

```{r pdtab, echo=FALSE, eval = FALSE}
i <- c(2, 3, 4, 13, 14)
pData(hl)[i, ]
```

@@

In addition to extracting new phenotypes, novelty detection methods 
are also useful for confirming the presence of known or postulated 
clusters in an unbiased fashion. For example, in [@hyper] 
the `phenoDisco` algorithm was used to confirm the data-specific
presence of the nucleus and nucleus sub-compartments. In the code 
chunk below, we demonstrate how to do this analysis, highlighting some of the 
optional arguments and parameters available for phenotype extraction 
and give some advice on how to interpret the output.

As the `phenoDisco` algorithm is semi-supervised it uses both 
labelled (markers) and unlabelled data to explore the data structure 
and find new sub-cellular data clusters. Thus the first step is to 
define some input labelled data i.e. markers, that the algorithm 
will use as input for the supervised learning aspect of the 
algorithm. As described in [@hyper] we define a set of markers 
to use as input for the analyses that cover well-known residents 
from three distinct organelle structures; the mitochondria, plasma 
membrane and ER, and from three well-known and abundant protein 
complexes; the proteasome and two ribosomal subunits, 40S and 60S. 
These input markers are stored in the `featureData` column of 
`hl` where `fcol = "phenoDisco.Input"`. We can use the convenience 
accessor function `getMarkers` to print out a table of the markers 
contained in this marker set. These initial markers were manually 
curated using information from the UniProt database, the Gene 
Ontology and the literature.

@@

In the code chunk below we demonstrate how to run the `phenoDisco` function and 
return a novelty detection result, according 
to the specified parameters. The algorithm parameters `times`, 
`GS` and `p` are passed to the function, along with the `fcol` 
to tell the algorithm where the input training data is contained.

```{r runPD, eval = FALSE}
## As per Christoforou et al (2016), 
hl <- phenoDisco(hl, fcol = "phenoDisco.Input", 
                 times = 200, 
                 GS = 20)
```

*Note: We do not evaluate this code chunk in this document as the algorithm is computational intensive and best parallelised over multiple workers. This phenoDisco analysis took ~24 hours to complete when parallelised over 40 workers.*

The argument `times` indicated the number of times we run unsupervied 
Gaussian Mixture Modelling before defining a new phenotype cluster. 
The recommended minimum and default value is 100. In the above 
code chunk we increase the value to `times = 200` as we have found 
for larger datasets (e.g. 5000+ proteins) a higher times is requried 
for convergence. `GS` defines the minimum number of proteins allowed 
per new data cluster and thus heavily influences what type of new 
clusters are extracted. For example, if a user is interesed in the 
detection of small complexes they may wish to use a small `GS = 10`, 
or `= 20` etc. If they wish to detect larger more abundant sub-cellular 
niches a much higher `GS` would be preferable. Specifying a small `GS` can be 
more time consuming than using a larger `GS`, and there is a trade off 
between finding interesting small complexes and those that may not 
be of interest as we find there is a tendancy to find more noise when 
using a small `GS` compared to using a higher one.

One may also consider increasing the search space for new data 
clusters by increasing the value of the parameter `G`. This defines 
the number of GMM components to test and fit; the default is `G = 1:9` 
(`mclust`'s default). One should note that the decreasing the 
`GS`, and increasing the values of the arguments `times`, `G` 
(among other function arguments, see `?phenoDisco`) will heavily 
influence (increase) the total time taken to run the algorithm. 
`phenoDisco` supports parallelisation and we strongly suggest 
you make use of a parallel processing to run these analyses (different 
backends can be set with argument `BPPARAM`).

```{r loadPDres, include=FALSE}
load("Data/pd_runs/pd.rda")
hl <- pdRes
setStockcol(NULL)
setStockcol(paste0(getStockcol(), 70))
```

The ouput of running the `phenoDisco` algorithm is a `MSnSet` containing 
the new data clusters, appended to the `featureData` under the name 
`pd`. We can see by typing `hl@processingData` directly into the console the 
processing information has been updated to the `MSnSet` recording the 
parameters that were used to run the analyses. This is handy for keeping 
track of data analyses. The results can be displayed by using the 
`getMarkers` function. We see that 5 new phenotype data clusters were 
found.

```{r showPDres}
hl
getMarkers(hl, fcol = "pd")
```

@@

We can also examine the each phenotype intercatively and visualise 
their protein profiles by using the `pRolocVis` function in the 
`pRolocGUI`. We found that phenotype 1 was enriched in nucleus associated proteins, phenotype 2 in chromatin associated proteins, phenotype 3 in cytosolic and phenotypes 4 and 5 in lysosomal and endosomal proteins.
